---
title: "Injury Analysis 2025"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

# Context

Due to a recent unfortunate accident in South America public concerns
about our company's workplace safety practicies have been voiced and are
requesting answers.

As a global industrial manufacturing company operating internationally,
ensuring compliance with local industrial regulations has split the
company into unique safety regimes, all of which are under the
microscope of the public eye.

The purpose of this report is to perform an analysis of workplace injury
data to help inform our response to recent journalist questions.

# Research Questions

To help inform the CEO's response to this crisis, she has provided data
on workplace injury within the company, as well as research questions to
guide the analysis.

1.  Of the various **safety regimes** in place across your company,
    which one would you recommend become the international standard for
    your company, based solely on injury prevention performance?

2.  It has been suggested by senior management that industry
    **experience** is more important than the safety regime when it
    comes to preventing injuries. His idea is that a policy should be
    developed that is directly related to lowering employee turnover
    will reduce injury rates. Does the available data support this
    assertion?

3.  Is there any relationship between:

-   Injuries and the annual **bonuses** a proportion of employees
    received\
-   Injuries and whether staff have received any formal external
    **qualifications** e.g. external safety training or a university
    degree.

# Summary of Available Data

The data provided for this report `injury.csv` contains counts of
`Injuries` and `Hours` worked for the past 12 months of operation,
aggregated by the `Experience` level of the workers and the workplace
`Safety` regime in place at their factory as well as other variables
that may have an affect on the number of `Injuries` at the factory.

More specifically:

-   `Injuries` – count of injuries in the group\
-   `Safety` – the safety regime in place for the group\
-   `Hours` – total hours worked by the group, e.g. if 2 workers worked
    10 hours, that's 20 hours total
-   `Experience` – the experience level of the group\
-   `bonus` – proportion of the group who received an annual bonus last
    year\
-   `training` – proportion of the group who have completed external
    safety training\
-   `university` – proportion of the group who have at least one
    university degree

It is also important to note that there is no provided information on
how many workers there are per group, or if all the groups have the same
amount of workers. Therefore, for the purposes of this report, it is
assumed that the number of workers in each group is insignificant to the
number of Injuries that happens within a `report_id`

# Data Processing / EDA

To inform the CEO with the relevant information to respond to journist's
questions, an Exploratory Data Analysis was conducted.

## Load data and libraries

```{r}
data <- read.csv("injury-3.csv")
```

```{r}
# code credit to: https://statsandr.com/blog/an-efficient-way-to-install-and-load-r-packages/

# Package names
packages <- c("GGally", "ggpubr", "tidyverse", "backports", "patchwork", "MASS")

# Install packages not yet installed
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Packages loading
invisible(lapply(packages, library, character.only = TRUE))
```

## Uni-variate Observations

Firstly, the raw tabular data was inspected to find any uni-variate
distributions of interest.

```{r}
nrow(data) # get this to use for histogram bins later
```

```{r}
vars <- names(data)

plots <- lapply(vars, function(v) {
  ggplot(data, aes(x = .data[[v]])) +
    geom_histogram(bins = (nrow(data) %/% 9)) # 72 / 8 = 9
})

wrap_plots(plots, nrow = 2)
```

```{r}
min(data$Hours)
max(data$Hours)
```

From these plots, the following observations can be made about the data:

-   `record_id` - as expected, are roughly sequential and uniform across
    the histogram, no obvious missing values can be seen.

-   `Injuries` - shows a strong right skewed distribution that also may
    be exponential, suggesting a poisson regression may be used to
    predict `Injuries`.

-   `Safety`- four discrete spikes at 1, 2, 3, 4 show that `Safety` is a
    **categorical factor** describing 4 different `Safety` regime.

-   `Experience` - also show discrete spikes at 1, 2, 3, 4 showing that
    `Experience` is also a **categorical factor** describing 4 different
    `Experience` levels, however, 2 and 3 are slightly under represented
    compared to 1 and 4.

-   `Hours` - also shows a strong right skewed distribution, and, using
    `min` & `max(data$Hours)` can be seen that the range is
    $3474-2135146$. With a similar uni-variate distribution to
    `Injury` - as well as `Hours` worked, intuitively being a factor
    that causes injury - this supports using an $\text{Injury Rate}$ per
    $\text{100 000}$ `Hours`, or, due to both following an exponential
    distribution, using the $\log{(\text{Injury Rate})}$ for further
    multivariate analysis and $\log(\text{Hours})$ as the offset
    function when fitting a model to the data.

-   `bonus`, `training` and `university` - either weak right or left
    skews can be observed in these graphs. They all appear to be roughly
    normally distributed from 0 - 1 and all are **continuous factors**,
    however nothing meaningful can be extracted from just uni-variate
    analysis and more analysis is required to understand if these
    factors play any role in understanding injury prevention.

With the uni-variate analysis, it can be seen that the most suited GLM
for modelling this data-set would be a Poisson Regression model. It can
also be seen that using
$\log{\frac{\text{Injuries}}{\text{Hours}}}\cdot\text{100 000}$ to model
this Poisson Regression model could be used for easier readability.

## Multi-variate Observations

With the uni-variate observations done,
$\log{\frac{\text{Injuries}}{\text{Hours}}}\cdot\text{100 000}$ will now
be used to conduct a multi-variate analysis to find significant factors.

`boxplots` can be used for the categorical, `Safety` and `Experience`
variables in the data-set, whereas `bonus`, `training` and `university`
are indiscreet, continuous variables requiring scatter-plots for
multivariate analysis. All the variables will be plotted against the
$\log$ injury rate.

### Categorical Variables

```{r, warning=FALSE}
data$Safety <- as.factor(data$Safety)

p1 <- ggplot(data, aes(x = Safety, y = log(Injuries / Hours * 100000))) +
  geom_boxplot() +
  labs(x = "Safety", y = "log(Injury Rate)")

p2 <- ggplot(data, aes(x = Safety, y = Injuries)) +
  geom_boxplot() +
  labs(x = "Safety", y = "Injuries")

p1 + p2
```

From the above $\log(\text{Injury Rate})$ & `Injuries` vs `Safety` plots
it can be seen that, while more total Injuries have occurred when
`Safety` regime 3 is in-place, it doesn't have a significantly higher
$\log(\text{Injury Rate})$ meaning that more observation is required
when looking at the model to determine whether or not `Safety` regime is
relevant.

```{r, warning=FALSE}
data$Experience <- as.factor(data$Experience)

p1 <- ggplot(data, aes(x = Experience, y = log(Injuries / Hours * 100000))) +
  geom_boxplot() +
  labs(x = "Experience", y = "log(Injury Rate)")

p2 <- ggplot(data, aes(x = Experience, y = Injuries)) +
  geom_boxplot() +
  labs(x = "Experience", y = "Injuries")

p1 + p2
```

From the above $\log(\text{Injury Rate})$ & `Injuries` vs `Experience`
plots it can be seen that, there is a very strong negative correlation
between `Experience` level and the $\log(\text{Injury Rate})$,
indicating that a large amount of the variance in the
$\text{Injury Rate}$ can be explained with the `Experience` of the
factory workers, supporting senior management's suggestion that industry
`Experience` is important for preventing `Injuries` , however, more
research is required to say whether `Safety` regime is not useful, as
senior management also suggests.

### Continuous Variables

```{r}
p1 <- ggplot(data, aes(x = bonus, y = log(Injuries / Hours * 100000))) +
  geom_point() +
  labs(x = "bonus", y = "log(Injury Rate)")

p2 <- ggplot(data, aes(x = training, y = log(Injuries / Hours * 100000))) +
  geom_point() +
  labs(x = "training", y = "log(Injury Rate)")
  
p3 <- ggplot(data, aes(x = university, y = log(Injuries / Hours * 100000))) +
  geom_point() +
  labs(x = "university", y = "log(Injury Rate)")

p1 + p2 /
p3
```

The continuous variables don't seem to have much meaningful relation to
the $\log(\text{Injury Rate})$ when directly plotted against each other,
more observation is required when modelling to determine if the
variables `bonus`, `training` and `university` have any compounded
affect on the $\log(\text{Injury Rate})$.

# Modelling Approach and Justification

## Poisson Regression

$$
p(Y=y|\lambda)=\frac{\lambda^y\exp{(-\lambda)}}{y!},\text{for }y=0,1,\cdots
$$

As previously understood, the number of `Injuries` in a `record_id`
follow an exponential function, alongside `Hours` following a similar
distribution, these facts support using a Poisson regression algorithm
with the $\log{(\text{Hours})}$ for the offset function.

### Starting with a basic model with all the main effects

Firstly, a basic model containing all the main effects of the dataset.

```{r}
fit <- glm(
  Injuries ~ offset(log(Hours)) + Safety + Experience + bonus + training + university,
  family = "poisson",
  data = data
)
summary(fit)
```

Unexpectedly, the model appears to have pulled reason from the chaos in
the variables `bonus` and `training`, and, although much less
significant, the effect of `university` is only $7.88\%$ due to chance.
This shows that in an all encompassing model, these seemingly random
variables do have an effect on the final $\log{(\text{Injury Rate})}$
warranting further analysis.

### Let R decide which interactions are important

With the discoveries in the initial model, modelling the main effects,
`stepAIC` from `MASS` is used to choose a model by AIC in a Stepwise
Algorthm:

```{r}
step_fit <- stepAIC(
  fit,
  scope = ~ (Safety + Experience + bonus + training + university)^2,
  direction = "both"
)
```

### Compare models

```{r}
AIC(fit, step_fit)
BIC(fit, step_fit)
summary(step_fit)
```

While including 3-way interactions may reduce the AIC, such complexity
offers little practical value. Moreover, none of the potential 3-way
relationships in the data-set are theoretically meaningful, and their
inclusion would likely just result in over-fitting rather than insight.

# Validity of Model and Modelling Results

These results align with the previous relationships observed in the
Multi-variate analysis, however, interestingly, `Safety` regime 4
appears to have a statistically significant relationship causing a
higher $\log{(\text{Injury Rate})}$ with `Safety` regime's 1, 2 and 3
causing statistically insignificant differences in
$\log{(\text{Injury Rate})}$. Another expected observation from the
fitted model is the `Experience` values, as `Experience` level
increases, significant decreases in $\log{(\text{Injury Rate})}$ can be
seen from the increasingly negative coefficient estimates. The final
model's fit achieves a good $\text{AIC}$ score of $1064.0$ a
$\Delta-492.7$ improvement when compared to the previous, $1556.7$,
achieved by the base model.More explanation behind the statistical
significance of `bonus`, `training` and `university` is also uncovered
in the final model.

# Recommendations and Conclusions

## Answers to Research Questions

### 1. Recommended Safety Regime

Based on the final model, **Safety Regime 3** is linked with the lowest
$\log{(\text{Injury Rate})}$ with a coefficient of $\sim-3.05$ where
$-3.05<\min{\{0.00, -1.00, -1.02\}}$ . This safety regime proves to be
effective at using `bonuses` as incentives to reduce injury in high risk
jobs, as well as ensuring workers are propery trained and qualified for
high risk jobs. As such, it is recommended that **Safety Regime 3** is
adopted as an international standard for the company. Moreover, further
research into local regulations and potential legal constraints should
be conducted to ensure that the international implementation of this
Safety Regime complies with all applicable laws in each operating
region.

### 2. Importance of Industry Experience vs Safety Regime

The final model supports senior management's claims that Industry
Experience is a very important factor, where **the more experienced the
workers are, the less injuries occur**. However, `Safety` Regime still
contributes significantly to the $\log{(\text{Injury Rate})}$ within the
injury reports.

### 3. Influence of `bonuses`, `training` and `university`

#### `bonuses`

From the final model it can be seen that `bonuses`, by their-selves,
have a slight negative correlation with the $\log{(\text{Injury Rate})}$
showing that giving a worker a bonus, in and of itself reduces injuries.
However, two way interactions, such as `Experience:bonus`,
`Safety:bonus`, both produce positive correlations for predicting the
$\log{(\text{Injury Rate})}$ for all factors of `Experience` and
`Safety`. Logically this positive correlation makes sense as more
experienced workers would handle difficult, and possibly more risky jobs
and different safety regimes may use bonuses as incentives for those
more risky jobs. Linking back to `bonuses`'s coefficient is negative, it
shows that the incentives in these theoretically high risk jobs do, in
fact, work at reducing injuries in these jobs.

#### `training`

Alike `bonuses`, `training` also has a slight negative coefficient in
the final model where training proves effective at reducing injury in
factories. With the interaction between `training` and `Safety` it can
be seen that for regimes 2 and 4, `training` doesn't have much effect on
the $\log{(\text{Injury Rate})}$, however, for regime 3, a significantly
higher number of injuries occur as `training` increases. As this safety
regime is largely effective at reducing injury it can't be simply
assumed that the training in this regime is just ineffective, and this
could be due to this regime giving more training to those working in
higher risk jobs within the factory. More investigation is required to
determine if this is true or if the `training` within Safety Regime 3
needs improvement. As expected, the interaction between `training` and
`Experience` appears to go hand in hand at preventing injury within the
factories, the more experienced a worker is and the more training they
receive, the lower their chances are of injuring themselves. The
interaction between `training` and `bonus` was also deemed significant
by $\text{AIC}$ tests, showing that as these variables increase, the
rate of injury increases, this observation supports the theory that
`bonuses` are given to those in high risk jobs that require more
`training`.

#### `university`

The model follows intuition where the more qualified a group of workers
are, the less injuries occur with a slightly negative but significant
correlation between `university` and $\log{(\text{Injury Rate})}$. A
similar trend to `training` can be seen with the interactions between
`Safety` and `Experience` where safety regime 3 is giving higher risk
jobs to more qualified workers. Where the most experienced workers with
qualifications show a negative correlation with injury, lowering injury
rates with highly experienced and qualified workers.

## Conclusion

This report used a Poisson Regression Algorithm to fit a GLM to
workplace injury data, incorporating the whole dataset which includes:
the number of `Injuries` and total `Hours` worked by all workers in the
record, aggregated by the `Safety` regime of the factory and
`Experience` of the workers, including percentages for the proportion of
workers with `bonuses`, `training` and `university` degrees or
qualifications within the group. The final model achieved a strong fit
$\text{AIC : }1064.0$ confirming key relationships:

-   **Safety Regime 3** is most effective at injury prevention compared
    to other regimes in-place.

-   **More experience** significantly reduces injury rates.

-   Bonuses, training and university degrees when utilized correctly
    reduce injury rates.

A comprehensive injury reduction strategy should be put in action to
implement Safety Regime 3 internationally, retain experienced workers
and refine the usage of bonuses, training and implement education
programs to increase the proportion of qualified workers.
